---
title: 'Boston House Price Prediction'
author: "schetry2@illinois.edu"
date: "July 16, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(leaps)
library(mlbench)
library(ggplot2)
library(knitr)
library(kableExtra)

```



**Data Selection:**

From Library **mlbench**, I am using **BostonHousing2** dataset, which is a dataframe of 506 observations and 19 attributes. The project will consider **medv** variable as the response from its relationships with other predictor variables. The dataset has two categorical variables, viz, **town** and **chas** and would be evaluated for determining the response. Additionally, the **rad** variable is categorical with the values it has, and would be converted to a factor during the evaluation.

**Analysis:**

*a. Introduction:*

**Statement**: Here, I will make an attempt to analyse the dataset to select a right model for predicting house prices in Boston. 

**BostonHousing2** is a database with housing information around Boston city from the 1970 census, and lists down the median house prices. The dataframe **BostonHousing2** is the corrected version with additional spatial information and is available at: http://lib.stat.cmu.edu/datasets/

The original data are 506 observations on 14 variables, $medv\$ being the target variable:
The following table summarizes the attributes and their types:

| Attribute | Description                                                      | Data Type |
|-----------|------------------------------------------------------------------|-----------|
| **crim** | per capita crime rate by town                                         | Factor    |
| **zn** | proportion of residential land zoned for lots over 25,000 sq.ft         | num       |
| **indus** | proportion of non-retail business acres per town                     | num       |
| **chas** | Charles River dummy variable (= 1 if tract bounds river; 0 otherwise) | Factor    | 
| **nox** | nitric oxides concentration (parts per 10 million)                     | num       |
| **rm** | average number of rooms per dwelling                                    | num       |
| **age** | proportion of owner-occupied units built prior to 1940                 | num       |
| **dis** | weighted distances to five Boston employment centres                   | num       |
| **rad** | index of accessibility to radial highways                              | Factor*** |
| **tax** | full-value property-tax rate per USD 10,000                            | int       |        
| **ptratio** | pupil-teacher ratio by town                                        | num       |
| **b** | 1000(B - 0.63)^2 where B is the proportion of blacks by town             | num       |	        
| **lstat** | percentage of lower status of the population                         | num       |
| **medv** | median value of owner-occupied homes in USD 1000's                    | num       |
| **cmedv** | corrected median value of owner-occupied homes in USD 1000's         | num       |
| **town** | name of town                                                          | Factor    |
| **tract** | census tract                                                         | int       |
| **lon** | longitude of census tract                                              | num       |
| **lat** | latitude of census tract                                               | num       |

(***): The attribute is converted to factor with 9 levels.

With regards to predicting the house prices, the attributes that I think would be valuable are: **ptratio, town, rad, dis, rm, nox and crim**

b. Methods:

```{r}
#install.packages("mlbench")
library(mlbench)
data("BostonHousing2")
BostonHousing2$rad = as.factor(BostonHousing2$rad)
str(BostonHousing2)
head(BostonHousing2, 10)


```


