---
title: "Week 9 - Homework"
author: "Saurav Prem Kaushik Chetry"
date: ''
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---


```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## Exercise 1 (`longley` Macroeconomic Data)

The built-in dataset `longley` contains macroeconomic data for predicting employment. We will attempt to model the `Employed` variable.

```{r, eval = FALSE}
View(longley)
?longley
```

**(a)** What is the largest correlation between any pair of predictors in the dataset?
```{r}
(cor_long = cor(longley))
max(cor_long[cor_long != 1])
```
The largest correlation is between the predictors "Year" & "GNP": <tt>`r max(cor_long[cor_long != 1])`</tt>

**(b)** Fit a model with `Employed` as the response and the remaining variables as predictors. Calculate and report the variance inflation factor (VIF) for each of the predictors. Which variable has the largest VIF? Do any of the VIFs suggest multicollinearity?

```{r}
w09_m1 = lm(Employed ~ ., data = longley)


```

The variance inflation factor for each predictor are:
```{r warning=FALSE}
library(faraway)
vif(w09_m1)
which.max(vif(w09_m1))
```

The variable with largest VIF is: GNP.

Yes, GNP.deflator, GNP, Unemployed, Population and Year all have VIF > 5 and they indicate multicollinearity.

**(c)** What proportion of the observed variation in `Population` is explained by a linear relationship with the other predictors?
```{r}
w09_m2 = lm(Population ~ . -Employed, data = longley)
(summary(w09_m2)$r.squared)
```

The proportion of the observed variation in "Population" explained by a linear relationship with other predictors is: <tt>`r summary(w09_m2)$r.squared * 100`$\%$</tt>


**(d)** Calculate the partial correlation coefficient for `Population` and `Employed` **with the effects of the other predictors removed**.
```{r}
cor(resid(w09_m1), resid(w09_m2))
```

the partial correlation coefficient for `Population` and `Employed` is: <tt>`r cor(resid(w09_m1), resid(w09_m2))`</tt>.

**(e)** Fit a new model with `Employed` as the response and the predictors from the model in **(b)** that were significant. (Use $\alpha = 0.05$.) Calculate and report the variance inflation factor for each of the predictors. Which variable has the largest VIF? Do any of the VIFs suggest multicollinearity?

```{r}
summary(w09_m1)$coefficients[,4][summary(w09_m1)$coefficients[,4] < 0.05]
```

Based on $\alpha = 0.05$, the significant predictors are : Unemployed, Armed.Forces and Year
```{r}
w09_m3 = lm(Employed ~ Unemployed + Armed.Forces + Year, data = longley)

```

the variance inflation factor for each of the predictors:
```{r}
vif(w09_m3)
which.max(vif(w09_m3))
```

The variable with max VIF is: Year.

None of the predictors indicate multicollinearty.

**(f)** Use an $F$-test to compare the models in parts **(b)** and **(e)**. Report the following:

- The null hypothesis
- The test statistic
- The distribution of the test statistic under the null hypothesis
- The p-value
- A decision
- Which model you prefer, **(b)** or **(e)**

```{r}
anova(w09_m3, w09_m1)

```

The null hypothesis suggests there is no statistical significance of the predictors "GNP.deflator","Population" and "GNP" for deriving response Employed, meaning both models are statistically similar, i.e. 

$\beta\ _0\ _{GNP.deflator}$ = $\beta\ _0\ _{GNP}$ = $\beta\ _0\ _{Population}$ = 0.

The test statistic is :<tt>`r anova(w09_m3, w09_m1)[2,5] `</tt>

Under null hypothesis the test statistic follow F-distribution with DF1 = 3 & DF2 = 9

The $p-value$ is <tt>`r anova(w09_m3, w09_m1)[2,6]`</tt>

Decision: $p-value$ is higher than common significance levels for regression, therefore null hypothesis Fails to be Rejected. This means the predictors GNP.deflator, GNP and Population are not statistically significant.

Preferred model: Model of **(e)**, Employed ~ Unemployed + Armed.Forces + Year


**(g)** Check the assumptions of the model chosen in part **(f)**. Do any assumptions appear to be violated?

```{r message=FALSE, warning=FALSE, paged.print=FALSE}

plot_fitted_resid = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  plot(fitted(model), resid(model), 
       col = pointcol, pch = 20, cex = 1.5,
       xlab = "Fitted", ylab = "Residuals")
  abline(h = 0, col = linecol, lwd = 2)
}

plot_qq = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  qqnorm(resid(model), col = pointcol, pch = 20, cex = 1.5)
  qqline(resid(model), col = linecol, lwd = 2)
}

par(mfrow = c(1, 2))
plot_fitted_resid(w09_m3)
plot_qq(w09_m3)
library("lmtest")
bptest(w09_m3)
shapiro.test(resid(w09_m3))
```

From the plots it appears the constant variance assumptions is followed but the normal distribution looks uncertain. Performing BP test and Shakiro Wilk test, the results show that the $p-values$ from the tests are high to fail to reject null hypothesis. This shows both constant variance and normality assumptions are followed.

***

## Exercise 2 (`Credit` Data)

For this exercise, use the `Credit` data from the `ISLR` package. Use the following code to remove the `ID` variable which is not useful for modeling.

```{r warning=FALSE}
library(ISLR)
data(Credit)
Credit = subset(Credit, select = -c(ID))
#?Credit
```

Use `?Credit` to learn about this dataset.

**(a)** Find a "good" model for `balance` using the available predictors. Use any methods seen in class except transformations of the response. The model should:

- Reach a LOOCV-RMSE below `135`
- Obtain an adjusted $R^2$ above `0.90`
- Fail to reject the Breusch-Pagan test with an $\alpha$ of $0.01$
- Use fewer than 10 $\beta$ parameters

Store your model in a variable called `mod_a`. Run the two given chunks to verify your model meets the requested criteria. If you cannot find a model that meets all criteria, partial credit will be given for meeting at least some of the criteria.

```{r warning=FALSE}
#based on DU's suggestions during OH
library(leaps)
balance_mod = lm(Balance~ ., data = Credit)
sumary(balance_mod)
vif(balance_mod)
best_model = lm(Balance ~ log(Income) + Limit + Cards + Student + Age + Ethnicity + Married, data = Credit)
mod_a = best_model



```

```{r, message = FALSE, warning = FALSE}
library(lmtest)

get_bp_decision = function(model, alpha) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_sw_decision = function(model, alpha) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```

```{r}
get_loocv_rmse(mod_a)
get_adj_r2(mod_a)
#get_sw_decision(mod_b, alpha = 0.01)
get_bp_decision(mod_a, alpha = 0.01)
get_num_params(mod_a)
```

**(b)** Find another "good" model for `balance` using the available predictors. Use any methods seen in class except transformations of the response. The model should:

- Reach a LOOCV-RMSE below `125`
- Obtain an adjusted $R^2$ above `0.91`
- Fail to reject the Shapiro-Wilk test with an $\alpha$ of $0.01$
- Use fewer than 25 $\beta$ parameters

Store your model in a variable called `mod_b`. Run the two given chunks to verify your model meets the requested criteria. If you cannot find a model that meets all criteria, partial credit will be given for meeting at least some of the criteria.
```{r warning=FALSE}

#based on DU's OH, continuing with model from part a

best_model_b = lm(Balance ~ I(Income^2) + log(Limit) + Cards + Student + Student:Limit + Student:Income + Student:Cards + Age + Ethnicity + Married, data = Credit)
mod_b = best_model_b

```

  
```{r, message = FALSE, warning = FALSE}
library(lmtest)

get_bp_decision = function(model, alpha) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_sw_decision = function(model, alpha) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```

```{r}
get_loocv_rmse(mod_b)
get_adj_r2(mod_b)
get_sw_decision(mod_b, alpha = 0.01)
#get_bp_decision(mod_a, alpha = 0.01)
get_num_params(mod_b)
```

***

## Exercise 3 (`Sacramento` Housing Data)

For this exercise, use the `Sacramento` data from the `caret` package. Use the following code to perform some preprocessing of the data.
```{r eval=FALSE}
install.packages("caret")
?Sacramento

```


```{r warning=FALSE}
library(caret)
library(ggplot2)
data(Sacramento)
sac_data = Sacramento
sac_data$limits = factor(ifelse(sac_data$city == "SACRAMENTO", "in", "out"))
sac_data = subset(sac_data, select = -c(city, zip))
```

Instead of using the `city` or `zip` variables that exist in the dataset, we will simply create a variable (`limits`) indicating whether or not a house is technically within the city limits of Sacramento. (We do this because they would both be factor variables with a **large** number of levels. This is a choice that is made due to laziness, not necessarily because it is justified. Think about what issues these variables might cause.)

Use `?Sacramento` to learn more about this dataset.

A plot of longitude versus latitude gives us a sense of where the city limits are.

```{r}
qplot(y = longitude, x = latitude, data = sac_data,
      col = limits, main = "Sacramento City Limits ")
```

After these modifications, we test-train split the data.

```{r}
set.seed(420)
sac_trn_idx  = sample(nrow(sac_data), size = trunc(0.80 * nrow(sac_data)))
sac_trn_data = sac_data[sac_trn_idx, ]
sac_tst_data = sac_data[-sac_trn_idx, ]
```

The training data should be used for all model fitting. Our goal is to find a model that is useful for predicting home prices.

**(a)** Find a "good" model for `price`. Use any methods seen in class. The model should reach a LOOCV-RMSE below 77,500 in the training data. Do not use any transformations of the response variable.

```{r warning=FALSE}
#exhaustive search 
library(leaps)
price_mod = lm(price ~ . ^ 2, data = sac_trn_data)
price_mod_ini = lm(price ~ 1, data = sac_trn_data)
all_price_mod = summary(regsubsets(price ~ . ^ 2, data = sac_trn_data))
best_r2_ind = which.max(all_price_mod$adjr2)
all_price_mod$which[best_r2_ind,]
p = length(coef(price_mod))
n = length(resid(price_mod))
#price_mod_aic = n * log(all_price_mod$rss / n) + 2 * (2:p)
price_mod_bic = n * log(all_price_mod$rss / n) + log(n) * (2:p)

#best_aic_ind = which.min(price_mod_aic)
best_bic_ind = which.min(price_mod_bic)

#all_price_mod$which[best_aic_ind, ]
all_price_mod$which[best_bic_ind, ]

 price_mod_best_aic = lm(price ~ beds + longitude + limits + beds:sqft + beds:type + beds:longitude + sqft:latitude + longitude:limits  , data = sac_trn_data)
 price_mod_best_bic = lm(price ~ beds + longitude  +limits + beds:sqft + beds:longitude + sqft:latitude + longitude:limits, data = sac_trn_data)
# best_price_model = price_mod_best_aic
 best_price_model = price_mod_best_bic
 
 get_loocv_rmse(best_price_model)



```

LOOCV-RMSE is <tt>`r get_loocv_rmse(best_price_model)`</tt>

**(b)** Is a model that achieves a LOOCV-RMSE below 77,500 useful in this case? That is, is an average error of 77,500 low enough when predicting home prices? To further investigate, use the held-out test data and your model from part **(a)** to do two things:

- Calculate the average percent error:
\[
\frac{1}{n}\sum_i\frac{|\text{predicted}_i - \text{actual}_i|}{\text{predicted}_i} \times 100
\]
- Plot the predicted versus the actual values and add the line $y = x$.

Based on all of this information, argue whether or not this model is useful.

```{r warning=FALSE}
sac_pred = predict(best_price_model, newdata = sac_tst_data);
(avg_pct_err = mean(abs(sac_pred - sac_tst_data$price)/sac_pred ) * 100 ) 

qplot(sac_tst_data$price,sac_pred)+ xlim(0,700000) + ylim(0,700000) + 
  geom_abline(intercept = 0, slope = 1)
     

     
     
# plot(sac_pred, 
#      sac_tst_data$price, 
#      col = "dodgerblue", 
#      pch = 20,
#      main = "Prediction vs Actual",
#      xlab = "Prediction",
#      ylab = "Actual",
#      abline(line(sac_pred,sac_tst_data$price), col="darkorange", lwd = 3)
#      )

```

The average percent error is <tt>`r avg_pct_err`$\%$</tt>, therefore the model is correctly predicting almost 75% of the times. Also, the predicted and actual values are almost equally distributed about the diagonal line.

***

## Exercise 4 (Does It Work?)

In this exercise, we will investigate how well backwards AIC and BIC actually perform. For either to be "working" correctly, they should result in a low number of both **false positives** and **false negatives**. In model selection,

- **False Positive**, FP: Incorrectly including a variable in the model. Including a *non-significant* variable
- **False Negative**, FN: Incorrectly excluding a variable in the model. Excluding a *significant* variable

Consider the **true** model

\[
Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \beta_5 x_5 + \beta_6 x_6 + \beta_7 x_7 + \beta_8 x_8 + \beta_9 x_9 + \beta_{10} x_{10} + \epsilon
\]

where $\epsilon \sim N(0, \sigma^2 = 4)$. The true values of the $\beta$ parameters are given in the `R` code below.

```{r}
beta_0  = 1
beta_1  = -1
beta_2  = 2
beta_3  = -2
beta_4  = 1
beta_5  = 1
beta_6  = 0
beta_7  = 0
beta_8  = 0
beta_9  = 0
beta_10 = 0
sigma = 2
```

Then, as we have specified them, some variables are significant, and some are not. We store their names in `R` variables for use later.

```{r}
not_sig  = c("x_6", "x_7", "x_8", "x_9", "x_10")
signif = c("x_1", "x_2", "x_3", "x_4", "x_5")
```

We now simulate values for these `x` variables, which we will use throughout part **(a)**.

```{r}
set.seed(420)
n = 100
x_1  = runif(n, 0, 10)
x_2  = runif(n, 0, 10)
x_3  = runif(n, 0, 10)
x_4  = runif(n, 0, 10)
x_5  = runif(n, 0, 10)
x_6  = runif(n, 0, 10)
x_7  = runif(n, 0, 10)
x_8  = runif(n, 0, 10)
x_9  = runif(n, 0, 10)
x_10 = runif(n, 0, 10)
```

We then combine these into a data frame and simulate `y` according to the true model.

```{r}
sim_data_1 = data.frame(x_1, x_2, x_3, x_4, x_5, x_6, x_7, x_8, x_9, x_10,
  y = beta_0 + beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + 
      beta_5 * x_5 + rnorm(n, 0 , sigma)
)
```

We do a quick check to make sure everything looks correct.

```{r}
head(sim_data_1)
```

Now, we fit an incorrect model.

```{r}
fit = lm(y ~ x_1 + x_2 + x_6 + x_7, data = sim_data_1)
coef(fit)
```

Notice, we have coefficients for `x_1`, `x_2`, `x_6`, and `x_7`. This means that `x_6` and `x_7` are false positives, while `x_3`, `x_4`, and `x_5` are false negatives.

To detect the false negatives, use:

```{r}
# which are false negatives?
!(signif %in% names(coef(fit)))
```

To detect the false positives, use:

```{r}
# which are false positives?
names(coef(fit)) %in% not_sig
```

Note that in both cases, you could `sum()` the result to obtain the number of false negatives or positives.

**(a)** Set a seed equal to your birthday; then, using the given data for each `x` variable above in `sim_data_1`, simulate the response variable `y` 300 times. Each time,

- Fit an additive model using each of the `x` variables.
- Perform variable selection using backwards AIC.
- Perform variable selection using backwards BIC.
- Calculate and store the number of false negatives for the models chosen by AIC and BIC.
- Calculate and store the number of false positives for the models chosen by AIC and BIC.

Calculate the rate of false positives and negatives for both AIC and BIC. Compare the rates between the two methods. Arrange your results in a well formatted table.

```{r paged.print=FALSE}
set.seed(19850208)
num_sims = 300
false_negative_aic_sim1 = rep(0, num_sims)
false_negative_bic_sim1 = rep(0, num_sims)
false_positive_aic_sim1 = rep(0, num_sims)
false_positive_bic_sim1 = rep(0, num_sims)


for (i in 1:num_sims) {
sim_data_1$y = beta_0 +  beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + beta_5 * x_5 + rnorm(n, 0, sigma)
full_model = lm(y ~ ., data = sim_data_1)

len = length(resid(full_model))
aic_model = step(full_model, direction = "backward", trace = 0)
bic_model = step(full_model, direction = "backward", k = log(len), trace = 0)

false_negative_aic_sim1[i] = sum(!(signif %in% names(coef(aic_model))))
false_positive_aic_sim1[i] = sum(names(coef(aic_model)) %in% not_sig)
false_negative_bic_sim1[i] = sum(!(signif %in% names(coef(bic_model))))
false_positive_bic_sim1[i] = sum(names(coef(bic_model)) %in% not_sig)
}

library(knitr)
results_aic_bic = data.frame(
Category = c("AIC", "BIC"),
False_Negatives = c(
mean(false_negative_aic_sim1),
mean(false_negative_bic_sim1)
),
False_Positives = c(
mean(false_positive_aic_sim1),
mean(false_positive_bic_sim1)
)
)

kable(results_aic_bic, digits = 2)
```

BIC method has lesser False_Positives. Both AIC & BIC avoided False_Negatives. The reasons are explained in part **b**

**(b)** Set a seed equal to your birthday; then, using the given data for each `x` variable below in `sim_data_2`, simulate the response variable `y` 300 times. Each time,

- Fit an additive model using each of the `x` variables.
- Perform variable selection using backwards AIC.
- Perform variable selection using backwards BIC.
- Calculate and store the number of false negatives for the models chosen by AIC and BIC.
- Calculate and store the number of false positives for the models chosen by AIC and BIC.

Calculate the rate of false positives and negatives for both AIC and BIC. Compare the rates between the two methods. Arrange your results in a well formatted table. Also compare to your answers in part **(a)** and suggest a reason for any differences.

```{r}
set.seed(420)
x_1  = runif(n, 0, 10)
x_2  = runif(n, 0, 10)
x_3  = runif(n, 0, 10)
x_4  = runif(n, 0, 10)
x_5  = runif(n, 0, 10)
x_6  = runif(n, 0, 10)
x_7  = runif(n, 0, 10)
x_8  = x_1 + rnorm(n, 0, 0.1)
x_9  = x_1 + rnorm(n, 0, 0.1)
x_10 = x_2 + rnorm(n, 0, 0.1)

sim_data_2 = data.frame(x_1, x_2, x_3, x_4, x_5, x_6, x_7, x_8, x_9, x_10,
  y = beta_0 + beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + 
      beta_5 * x_5 + rnorm(n, 0 , sigma)
)
```

```{r}
set.seed(19850208)
num_sims = 300
false_negative_aic_sim2 = rep(0, num_sims)
false_negative_bic_sim2 = rep(0, num_sims)
false_positive_aic_sim2 = rep(0, num_sims)
false_positive_bic_sim2 = rep(0, num_sims)


for (i in 1:num_sims) {
sim_data_2$y = beta_0 +  beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + beta_5 * x_5 + rnorm(n, 0, sigma)
full_model_2 = lm(y ~ ., data = sim_data_2)

len_2 = length(resid(full_model_2))
aic_model_2 = step(full_model_2, direction = "backward", trace = 0)
bic_model_2 = step(full_model_2, direction = "backward", k = log(len_2), trace = 0)

false_negative_aic_sim2[i] = sum(!(signif %in% names(coef(aic_model_2))))
false_positive_aic_sim2[i] = sum(names(coef(aic_model_2)) %in% not_sig)
false_negative_bic_sim2[i] = sum(!(signif %in% names(coef(bic_model_2))))
false_positive_bic_sim2[i] = sum(names(coef(bic_model_2)) %in% not_sig)
}

library(knitr)
results_aic_bic_2 = data.frame(Category = c("AIC", "BIC"),
                               False_Negatives = c(mean(false_negative_aic_sim2),mean(false_negative_bic_sim2)),
                               False_Positives = c(mean(false_positive_aic_sim2),mean(false_positive_bic_sim2))
)

kable(results_aic_bic_2, digits = 2)
```

In sim_data2, the x_8, x_9 have a linear relationship with x_1. Also, x_10 has linear relationship with x_2. Because of these relationship with significant variables, x_8, x_9 and x_10 are implied as significant.

However, as the calculations of false negatives and false positives still make use of the pre-defined vectors(signif & non_sig), we see a difference in results in part **a** & **b**.

We saw no false negatives in part **a** because both AIC and BIC models did not leave out any signigicant variables.
We see false negatives in part **b** because AIC and BIC models still leave out x_8,x_9 and x_10, which are now significant due to their linear relationship.

As BIC chooses a smaller model over AIC, the chances of it excluding significant variables are more, hence BIC model has higher average false negative than AIC model.

Similarly, as AIC model has more predictors over BIC, chances are it will include more insignificant variables and hence have higher average false positives compared to BIC model.
